---
title: "Multivariate Visualization of Painters Style"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Multivariate Visualization of Painters Style}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, 
                      collapse=TRUE, R.options=list(digits=4),
                      fig.align = "center")
```

Vignette built using `heplots`, version `r packageDescription("heplots")[["Version"]]`,
`candisc`, version `r packageDescription("candisc")[["Version"]]` and `ggplot2`, version `r packageDescription("ggplot2")[["Version"]]`.


## Introduction

The `painters` dataset from the `MASS` package contains subjective assessments of 54 classical painters on four aesthetic qualities: Composition, Drawing, Colour, and Expression. These ratings, on a scale of 0-20, were made by the 18th century art critic Roger de Piles [-@DePiles1743], a notable member of the French Academy of Painting and Sculpture. The ratings reflect his judgment of the "artistic merit" of these painters on each of these dimensions.  His painters include the well-known Michelangelo, Tintoretto, Titian, Rembrandt, Rubens, Durer and others that art historians know well.

The 54 painters are classified into eight schools: Renaissance (A), Mannerist (B), Seicento (C), Venetian (D), Lombard (E), Sixteenth Century (F), Seventeenth Century (G), and French (H). This data was first analysed by @Davenport1972. Jolliffe [-@Jolliffe1986] 
used it to illustrate principal components analysis, and the data later featured in @VenablesRipley2002, who put it into the
`MASS` package.

### Painters schools: Going further

Other analysts largely treated the schools of painting used by de Piles as discrete categories, and this is the path I follow here.

But, for the record, it is possible to go further, by placing these schools in approximate chronological order and describing some of
their characteristics, as shown in @tbl-schools-table.

```{r}
#| label: tbl-schools-table
#| echo: false
#| tbl-cap: "Characteristics of painting schools"
library(gt)
schools.tab <- data.frame(
  School = c("Renaissance", "Venetian", "Lombard", "16th C", 
             "Mannerist", "Sciento", "17th C", "French"),
  Period = c("1400-1520", "1450-1600", "1490-1600", "1500-1600",
             "1520-1600", "1600-1700", "1600-1700", "1600-1750"),
  Characteristics = c(
    "Emphasis on naturalism, perspective, classical ideals; focus on balance, harmony, and proportion. Key figures: Leonardo, Michelangelo, Raphael.",
    "Rich color and light effects (overlaps Renaissance/Mannerist periods); emphasis on sensuous surfaces, atmospheric effects. e.g.: Titian, Veronese, Tintoretto.",
    "Northern Italian style; naturalism with dramatic chiaroscuro; influenced by both Renaissance ideals and Venetian colorism. e.g.: Albani, Caravaggio",
    "Late Renaissance through Mannerism; transition period showing diverse regional styles across Europe.",
    "Reaction to Renaissance harmony; elongated figures, complex compositions, artificial elegance, and emotional intensity. e.g.,: Parmigiano",
    "Likely refers to 17th century Italian schools emphasizing dramatic realism and scientific observation of nature (possibly related to Caravaggio's influence).",
    "Baroque period; dramatic use of light and shadow, movement, emotional intensity, grandeur. Includes Caravaggism and various national Baroque styles.",
    "French Baroque and emerging Rococo; emphasis on classical restraint, academic standards, royal patronage. Poussin, Claude Lorrain, later Watteau."
  )
)

schools.tab |>
  gt() |>
  cols_width(
    School ~ px(80),
    Period ~ px(100),
    Characteristics ~ px(400)
  ) |>
  tab_header(title = "Chronological overview of painting schools") |>
  tab_options(column_labels.font.weight = "bold")
  tab_options(table.width = pct(100))
```

For example, for other analyses, the `School` variable could be made an ordered factor, or schools could be
categorized into in broad periods ("Early", "Transition", "Baroque"), or by main emphasis ("Form", "Color", "Drama")

### This vignette

This vignette demonstrates the use of multivariate visualization methods of the `candisc` package, and its' sister, `heplots`.
The main goal is to illustrate how
hypothesis-error (HE) plots, canonical discriminant analysis, and discriminant analysis plots help understand how the schools differ in their aesthetic characteristics. It shows that, while simple univariate and bivariate plots
are often useful, dimension reduction methods can give a clearer view.

Let's get started. Load packages, and the `painters` data.

```{r setup, message=FALSE}
library(MASS)
library(heplots)
library(candisc)
library(ggplot2)
library(car)
library(dplyr)
library(RColorBrewer)

data(painters, package = "MASS")
str(painters)
```

## Data Preparation

The `School` factor in the dataset just uses letters, "A" through "H".
We begin by using longer, more descriptive labels to identify the schools, making the plots easier to interpret.

```{r data-prep}
school <- c("Renaissance", "Mannerist", "Seiento", "Venetian",
            "Lombard", "16th C", "17th C", "French")
levels(painters$School) <- school
head(painters)
```

## How many from each school?

The distribution of painters across schools is unbalanced, with some schools having many more representatives than others.
In particular, the small $n = 4$ for 16th C and French schools
affect some multivariate methods (e.g., Box's M test for equality, `heplots::boxM()`)
which rely on estimating the within-group covariance matrices.

```{r school-counts}
table(painters$School)
```

## Exploratory plots

Before conducting formal multivariate analyses, we explore the data visually. 

First, set up colors and other graphic attributes to be consistent across plots. I chose the `RColorBrewer` discrete palette,
`Dark2`. 
```{r colors}
options(
  ggplot2.discrete.colour = function() scale_colour_brewer(palette = "Dark2"),
  ggplot2.discrete.fill = function() scale_fill_brewer(palette = "Dark2")
)

# base R graphics
school.colors <- brewer.pal(n = length(levels(painters$School)), name = "Dark2")
school.pch <- c(16, 17, 15, 9, 7, 8, 10, 5)
```

A boxplot shows the distribution of Colour scores across schools. You could do the same for the other variables.
In general, I try to suppress legends where possible, because the space they take reduces resolution for the data.

```{r boxplot, fig.width=8, fig.height=5}
ggplot(data = painters, aes(x = School, y = Colour, fill = School)) +
  geom_boxplot() +
  labs(title = "Colour Scores Distribution by Painting School",
       x = "School",
       y = "Colour Score (0-20)") +
  theme(legend.position = "none")
```

The Venetian and 17th C painters were given high marks for their use of color.

Alternatively, to examine all four aesthetic variables simultaneously, you need to reshape the data to long format.
Here, I create violin plots overlaid with individual points.

```{r violin-plots, fig.width=8, fig.height=5}
painters_long <- painters |>
  tidyr::pivot_longer(cols = c(Composition, Drawing, Colour, Expression),
                      names_to = "Metric", values_to = "Score")

ggplot(painters_long, aes(x = Metric, y = Score, fill = Metric)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.1) +
  labs(title = "Distribution of de Piles' Scores",
       y = "Score (0-20)") +
  theme_light(base_size = 15) +
  theme(legend.position = "none")
```

Note that de Piles used the full range of 0--20 for most of these measures, but the range for `Drawing` had no ratings in
the lower quarter.

## Sample scatterplots

Scatterplots of pairs of variables, with data ellipses for each school, can reveal the multivariate relationships and potential group separation, but they are limited to 2D views, where the other variables vary as they do.

<!-- When I map `shape` to `School`, `ggplot` complains because it only provides 6 shapes, so it is necessary to use -->
<!-- `scale_shape_manual()`. I avoid using a legend by labeling the schools at their means. -->

Data ellipses and labels for the schools at their group mean help to make such plots more understandable.
```{r scatter1, fig.width=8, fig.height=6}
means <- painters |>
  group_by(School) |>
  summarise(across(Composition:Expression, mean))

ggplot(painters,
       aes(Composition, Colour,
           color = School, shape = School)) +
  geom_point(size=3) +
  stat_ellipse(level = 0.68, linewidth = 1.3) +
  geom_label(data = means,
             aes(label = School)) +
  scale_shape_manual(values = school.pch) +
  theme_classic(base_size = 15) +
  theme(legend.position = "none")
```

You can see, for example,  that the Venetian school and 17th C painters were rated highest on use of color.

Simmilarly, here's a plot for Drawing and Expression:

```{r scatter2, fig.width=8, fig.height=6}
ggplot(painters,
       aes(Drawing, Expression,
           color = School, shape = School)) +
  geom_point(size=3) +
  stat_ellipse(level = 0.68, linewidth = 1.3) +
  geom_label(data = means,
             aes(label = School)) +
  scale_shape_manual(values = school.pch) +
  theme_classic(base_size = 15) +
  theme(legend.position = "none")
```

However, with so many schools, it is hard to get a clear idea of how they differ on these aesthetic qualities from bivariate plots.
Here's where multivariate methods can help.

## MANOVA: Testing school differences on the aesthetic qualities

We fit a multivariate linear model (MANOVA) to test whether schools differ significantly on the four aesthetic variables considered jointly. The model treats School as the predictor and the four aesthetic scores as a multivariate response.

```{r manova}
painters.mod <- lm(cbind(Composition, Drawing, Colour, Expression) ~ School, 
                   data = painters)
car::Anova(painters.mod)
```

With `School` as an unordered factor, the intercept represents
`Renaissance` painters. Coefficients for the other schools
represent their difference from Renaissance.
```{r coef}
coef(painters.mod)
```

## Check for multivariate outliers

Before interpreting the MANOVA results, we check for multivariate outliers using a chi-square Q-Q plot of squared Mahalanobis distances.
This looks OK, because all the painters are well within the confidence bounds.

```{r outliers, fig.width=6, fig.height=6}
cqplot(painters.mod, id.n = 3)
```

## HE plots

Hypothesis-error (HE) plots provide a visual representation of multivariate hypothesis tests. The **H** ellipse represents the hypothesis sum of squares and cross-products, while the **E** ellipse represents error (within-group) variation. When **H** extends anywhere outside **E**, there is evidence of a significant effect.

By default, the `heplot()` shows the first two variables (Composition and Drawing).

```{r heplot1, fig.width=7, fig.height=7}
heplot(painters.mod,
       fill = TRUE, fill.alpha = c(0.1, 0.05),
       cex.lab = 1.25)
```

This tells us that the means for the schools of painting are positively correlated on these variables.
The within-group residuals are also positively correlated, to about the same extent.

We can also examine other pairs of variables, such as `Colour` and `Expression`.

```{r heplot2, fig.width=7, fig.height=7}
heplot(painters.mod,
       variables = 3:4,
       fill = TRUE, fill.alpha = c(0.1, 0.05),
       cex.lab = 1.25)
```

Here, the means for the schools are negatively correlated, but what seems to drive this is large difference between the French
school, very high on Expression, but low on use of Colour, vs. the Venetian school which shows the reverse pattern.


## All pairwise HE plots

To get a comprehensive view of how schools differ across all pairs of variables, we create a matrix of pairwise HE plots,
provided by the `heplots::pairs.mlm()` method.

```{r pairs-heplot, fig.width=10, fig.height=10}
pairs(painters.mod)
```

The configuration of the means for the schools on these pairs of ratings might give food for thought to an art historian.

## Canonical analysis


Canonical discriminant analysis reduces the dimensionality of the multivariate differences among schools. It finds linear combinations of the original variables that maximize between-group variation relative to within-group variation.

```{r candisc}
painters.can <- candisc(painters.mod)
painters.can
```

The output shows how many dimensions (canonical variates) are needed to account for the differences among schools, along with the percentage of between-group variance explained by each dimension. The first two dimensions account for 81% of between-school
variance, but the third dimension is also significant, adding 15.8%.

## HE plot in canonical space

Plotting the data in canonical space provides the clearest view of group separation, as the canonical dimensions are specifically constructed to maximize differences among schools. Variable vector in this plot have components proportional to the
correlations of the observed aesthetic variables with the canonical scores.

```{r heplot-can, fig.width=8, fig.height=8}
heplot(painters.can,
       fill = TRUE, fill.alpha = c(0.1, 0.05),
       var.lwd = 2, var.cex = 1.4,
       cex.lab = 1.25,
       prefix = "Canonical dimension")
```

## 3D view

The analysis suggests three significant dimensions of differences among schools. We can visualize all three dimensions simultaneously using a 3D HE plot. This plot is not yet rendered in the vignette.

```{r heplot3d, eval=FALSE}
if(requireNamespace("rgl")) {
  heplot3d(painters.can, col = c("pink", "brown"))
}
```

## Discriminant analysis

Linear discriminant analysis (LDA) provides another approach to understanding group differences and can be used for classification. Unlike canonical discriminant analysis, which focuses on visualization for the MANOVA model, LDA is optimized for predicting group membership. But the `plot_discrim()` provides something similar, and the dimensions are essentially the same.

```{r lda}
painters.lda <- lda(School ~ .,
                    data = painters)
painters.lda
```

Again, we see that there are three dimensions that seem important in understanding how the Schools differ
in terms of the ratings on these four aesthetic attributes.

<!--
We can examine the confusion matrix to see how well the discriminant function classifies painters into their actual schools:

```{r confusion}
class_table <- table(
  actual = painters$School,
  predicted = predict(painters.lda)$class 
)
class_table
```
-->

The `plot_discrim()` function visualizes the discriminant analysis results by plotting observations in the space defined by the discriminant dimensions (LD1, LD2, etc.). Here we plot the first two discriminant dimensions:

```{r plot-discrim1, fig.width=9, fig.height=7}
plot_discrim(painters.lda, LD2 ~ LD1,
             labels = TRUE,
             labels.args = list(geom = "label")) +
  scale_shape_manual(values = school.pch) +
  theme_classic(base_size = 15) +
  theme(legend.position = "none")
```

We can also examine other pairs of discriminant dimensions, such as LD1 and LD3:

```{r plot-discrim2, fig.width=9, fig.height=7}
plot_discrim(painters.lda, LD3 ~ LD1,
             labels = TRUE,
             labels.args = list(geom = "label")) +
  scale_shape_manual(values = school.pch) +
  theme_classic(base_size = 15) +
  theme(legend.position = "none")
```

To complete this, it would be useful to add vectors for the variables in these plots as in the plots for the `painters.can` object shown above. This is not yet implemented in the package, but the vectors are given by columns of `cor_lda()`

```{r}
cor_lda(painters.lda)
```


## Summary

This vignette has demonstrated the use of HE plots, canonical discriminant analysis, and discriminant analysis plots to understand how classical painting schools differ in their aesthetic characteristics. HE plots provide intuitive visual tests of multivariate hypotheses, canonical analysis reveals the dimensions of maximum group separation, and discriminant analysis plots show how well groups can be distinguished and classified in reduced dimensional space.

## References
